Linking watershed nutrient loading to estuary water quality with semiparametric models

### 2023-08-07 Rebuttal_Letter/Rebuttal_Letter.Rmd

---
author: Michael Schramm
return-address: 
  - Texas A&M AgriLife
  - Texas Water Resources Institute
  - 1001 Holleman Drive E.
  - College Station, Texas 77840
return-short: Michael Schramm, 2118 TAMU, College Station, TX 77840
address:
  - Dr. James Latimer
  - Academic Editor
  - PeerJ Life & Environment

subject: Revised Manuscript
opening: Dear Dr. Latimer,
closing: Best regards,
signature: Michael Schramm

output: komaletter::komaletter
---

Thank you for the prompt and thorough review of my submitted manuscript, "Linking watershed nutrient loading to estuary water quality with semiparametric models." I really appreciate the substantive and helpful comments from all the reviewers. Line by line comments and responses are included below (reviewer comments are italicized).

**Reviewer 1**

*Very well-structured abstract – introduced the problem, the objectives of the research, the results obtained, and the conclusion. No changes are recommended.*
*The introduction was well-structured and easy to follow. No changes are recommended.*

**Response:** Noted, and thank you.

*Line 114: A link to the USGS NWIS system could be added.*

**Response:** The url has been added.

*A brief description of the sample collection and laboratory analyses described in the QAPP and TCEQ procedures manual, detection and recovery limits, and instrumentation would seem appropriate.*

**Response:** I added the EPA methods and TCEQ's required AWRL's to Tables 1 and 2. I believe this is sufficient documentation.

*Also, what are the state-recommended guideline values? Include them in Table 2*

**Response:** I've added the associated state water quality criteria or screening levels in both Tables 1 and 2.

*Line 371 “has” repeated twice*

**Response:** Thank you, this is fixed.

*The author adequately answered the research questions and fulfilled all the stated objectives. The results from the GAM model provided estimates of watershed nutrient loadings and an outlook of nutrient loading's influence on water quality.*
*The whole of the paper was well structured, easy to read, and contributes sufficiently to the existing body of knowledge on estuary anthropogenic pollution and water quality monitoring.*
*Through this paper, the author has created an addition to the body of work on using models for the prediction and assessment of pollution in estuaries and aquatic bodies, as a means of recommending and establishing policy actions for pollution mitigation.*

**Response:** Thank you for the review.

**Reviewer 3**

*Table 1: It would be helpful if you would include the watershed or station name under the USGS station ID in this table.*

**Response:** I've added the associated watersheds to each USGS station in Table 1.

*Line 164: Not sure if this is a typo, or if I’m missing what log1p(Q) is.*

**Response:** This was shorthand for *log(1+Q)*. I've changed this to *log(1+Q)* to be consistent with equation 1.

*Line 392, "Limitation" section: I expected this to be a discussion of limitations of the approach like matching daily loads to concentrations, the complex environmental factors that could be involved in chlorophyll and DO trends, changing climate, etc. However, since you focus on the (very valid) problem of limited sampling, you may want to rename this sub-section of the Discussion. Maybe call it "Data Needs" or "Suggested Management Focus" or something like that?*

*Line 53-54: Since only one of the trends you mention from Bugica is for a nitrogen species, I suggest rephrasing or expanding this sentence to explain why we’d be concerned about all of the parameters that are increasing (or clarify if we are only concerned about the TKN).*

*Lines 129-131: It would be helpful if you described how deep the water column is at the estuary sample locations and from what depth the samples that you analyzed were taken. A little more explanation of the estuary dynamics would be helpful too -- is this a stratified or well-mixed system, what is the range in salinity, is there a key season of concern for eutrophication?*

**Response:** I have added additional description of the estuary dynamics *(lines 101-108)*. Specifically mixing, turbidity, and salinity. Given the preliminary nature of this study, there is not a season of concern identified yet, so it is not included. *still need to add sample depth here*

*Lines 194-201: Some discussion is needed of how you dealt with temporal matching of flow and loads to concentrations in the estuary. I see in the discussion (line 373-374) that you matched flow and load estimated for the day of each sample. This approach should be explained in the methods, and I think you need some explanation is needed as to why that would be appropriate and/or how it may impact the results. As you discuss below, you might get better relationships (for Table 4, Lines 244-248) if you either lagged the match or matched an average load and flow from some reasonable period before the estuary sample date. If you had the ability to try some models like that, I think it may be a useful addition.*

*Line 296: Would damming of the Navidad River impact the loads through the Lavaca River watershed? I don’t see why if the data was from the USGS-08164000 site.*

**Response:** You are absolutely correct. I mistakenly wrote this section considering the
combined watershed loads, when these estimates were for the Lavaca River watershed only.
I've removed this and incorporated your statement below. Thank you very much for catching this!

*Lines 286-321: I think this discussion is too long comparing the TP yield first to the Dunn study, and then to 3 other studies. The TP yield estimate from your study is in the middle of the others, so you can condense this 2 paragraph discussion and simply say that you fall in the middle of other estimates and state briefly that there are reasons such as land use and flow variability that the estimates vary.*

**Response:** I agree, this section has been considerably condensed.

*Lines 360-361: It is probably worthwhile to note that the effect of flow you see in chlorophyll is very likely due to more nutrients coming in with high flows. I wouldn’t want a reader to think that since your flow-adjusted nutrient loads are not linked to chlorophyll that this means there is definitely no impact of nutrients on chlorophyll concentrations. Along those lines, I’d suggest in Line 350-251 you specific that “No relationships between flow-adjusted inorganic nitrogen or TP loadings with chlorophyll-a were observed”*

**Reviewer: Marcus Beck**

*I have some concerns about the methods starting on line 184 that describe how the load estimates are related to the estuarine data. First, I strongly encourage the authors to develop these models using cumulative or lagged estimates for the flow and load variables. It appears that these are currently modeled as the estimates for the same day the water quality parameter was sampled. I would not expect much of a relationship between water quality and the independent variables at that time scale. I suspect that the relationships the author shows regarding flow/load with the response reflects some level of temporal autocorrelation between the daily estimates and preceding conditions/loads, where the latter may be a truly causative relationship. I see two potential approaches to remedy the issue. One, the author could test different lagged values for flow and load, although this would require a bit of trial and error. A potentially better solution is to use the aggregate flow volume and load between the sample dates as a sum of the volume and mass, respectively, between each observation of the time series for the response variable. I would expect these aggregate measurements to be much more strongly correlated to the response variables than the daily values that are currently used. This limitation is noted on lines 371-374, so I’m curious why the authors did not use this approach.*

*My second concern about the models in eqns. 5-7 also relates to the flow and load variables. It’s stated on line 199-201 that flow-adjusted load values were used, which makes sense, but I don’t understand how this additional analysis differs from the flow-normalization in the previous section (lines 176 – 179). Why not just use the flow-adjusted loads from the previous models? Once this is clarified (or corrected), it might make more sense to change eqn 6 to include smoothers for ddate, yday, and flow-normalized load, and eqn 7 to included smoothers for ddate, yday, and load. These wouldn’t be “nested” models for AIC comparison, but would perhaps provide a better approach to evaluate loads related to natural or non-natural variation. I don’t see the value in including flow as a separate variable since it’s basically a surrogate for total load.*

*I also strongly encourage the author to consider replacing “semiparametric models” with “Generalized Additive Models” in the title. It’s a more accurate description of what’s included in the paper.*

*Line 18: Maybe replace “non-natural variation” with “anthropogenic sources”.*

**Response:** Suggested changed made.

*Line 61: Accidental line break. *

**Response:** Fixed.

*Line 77: Suggest rephrasing as “…generalized linear models where the response variable is modeled as the sum of…”*

**Response:** Suggested change made.

*Tables 1 and 2: It would be useful to indicate the approximate sampling interval for the water quality parameters. It looks like quarterly. *

*Line 102: Some more information about the Bay would be helpful. What is the average depth? Any information on residence time? Is the water column mixed, stratified, or seasonally-stratified?*

**Response:** I have not identified any information on residence time. I have added more information on average depth, mixing and stratification of the water column. **add line numbers**

*Lines 128 – 129: The interpretation of the figures, tables, and text would be much easier if the estuarine sites were simply referred to as “upper”, “mid”, and “lower”. I suggest replacing all station names with these terms throughout the manuscript. *

*Line 129 – 131: Some additional information about the sampling methods for the estuarine sites could be useful. Where in the water column were these samples taken? Is the water column mixed? I’m thinking specifically about dissolved oxygen. The results show that it was not related to load, but you may have different outcomes depending on surface vs bottom DO. *

*Line 138: Typo “tend”.*

**Response** Fixed typo.

*Figure 1: Remove “contribution” from the legend. *

**Response:** Suggested change made.

*Eqn 2: I think some additional details are needed in the text below this equation. Why was d set at 0.95? Based on the text, this seems awfully close to 1 or the “mean flow”. Also, the equation includes a lag term, t – t_j, but the size of this lag is not stated, i.e., “some historical day of observation”. Was this set explicitly by the author, and if so, is there a justification? *

*Line 151: It’s not clear to me how fa is a unitless term. The equations show these as simple differences of average streamflow, so wouldn’t the units be in log-cfs? *

*Line 169: What percentage of the observations for each parameter were non-detects? I realize that setting the values as ½ the limit can introduce bias, but this may not be a big deal if there aren’t many non-detects. There does appear to be a GAM package that handles non-detects with Tobit models, but it is quite old (by software standards). It may be worth exploring. https://cran.r-project.org/web/packages/cenGAM/index.html*

*Line 236: This might be a personal preference, but “unusual periodic pattern” seems like regular periodicity and not a more random periodicity as one would expect with rain events. Consider rephrasing this text. *

*Figure 2 caption: Consider adding a sentence like “Values closer to 1 for NSE and r and values closer to 0 for PBIAS represent more ideal goodness-of-fit assessments.”*

*Figure 3 caption: Consider adding Lavaca and Navidad with the station name, as in the caption for figure 4. *

*Figure 5 caption: Consider including a reference to Table 3, as it wasn’t immediately clear to me that these models didn’t include flow or load. *

*Table 4: As for my previous comment, I’d replace the site names with “upper”, “mid”, and “lower” and arrange them accordingly. The current arrangement is “mid”, “lower”, and “upper”. Also, it looks like several of the statistical summaries are identical, e.g. the inflow and inflow+load model for TP at TCEQ-13384. Are these simply rounding issues? And if so, are these models really all that different in terms of explained variance? *

*Line 272: The text “contributing tributaries” seems redundant. Maybe remove “contributing”. *

*Line 341: Replace “were” with “was”. *

*Line 395: Replace “suggest” with “such as”.*

*Line 420: Replace “provide” with “prove” or “be”.*

### 2023-08-07 Schramm-2023-06-PeerJ/Schramm-2023-08-PeerJ.Rmd
start working on revisions

---
title: Linking watershed nutrient loading to estuary water quality with semiparametric models
preprint: false
author: 
  - name: Michael Schramm
    affiliation: 1
    corresponding: true
    email: michael.schramm@ag.tamu.edu
affiliation:
  - code: 1
    address: Texas Water Resources Institute, Texas A\&M AgriLife Research, College Station, Texas, United States
abstract: >
  Evaluating estuary water quality responses to reductions (or increases) in nutrient loading attributed to on the ground management actions can be challenging due to the strong influence of environmental drivers on nutrient loads and non-linear relationships. This study applied semiparametric approaches to calculate watershed nutrient loads and assess responses in estuary water quality to seasonally-adjusted freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas. Lavaca Bay is a secondary embayment on the Texas coast displaying early potential for eutrophication and water quality degradation. Use of flow-adjusted nutrient loads allowed the study to evaluate the response in water quality to changes in nutrient loads driven by non-natural variation. Cross-validation indicated that, despite data constraints, semiparametric models performed well at nutrient load prediction. Based on these models, delivered annual nutrient loads varied substantially from year to year. In contrast, minimal changes in flow-normalized loads indicate that nutrient loadings were driven by natural variation in precipitation and runoff as opposed to changes in management of nonpoint sources. Models indicated no evidence of long-term changes in dissolved oxygen or chlorophyll-*a* within Lavaca Bay. However, site specific long-term increases in both organic and inorganic nitrogen are concerning for their potential to fuel eutrophication. Further analysis found freshwater inflow had strong influences on nutrient and chlorophyll-*a* concentrations but there was no evidence that changes in watershed nutrient loading explained additional variation in dissolved oxygen or chlorophyll-*a* concentrations. In addition to providing a baseline assessment of watershed nutrient loading and water quality responses in the Lavaca Bay watershed, this study provides methodological support for the use of semiparametric models in load regression models and estuary assessments.
bibliography: bibliography.bib
output:
  rticles::peerj_article:
    citation_package: biblatex
biblio-style: apa
header-includes:
    - \usepackage{csquotes}

@@

store <- "C:/Data-Analysis-Projects/lavaca-nutrients/_targets"
```

@@

While WRTDS is a statistical approach developed specifically for water quality applications, 
GAMs are a broadly applicable statistical method. GAMs are a semiparametric 
extension of generalized linear models where the linear predictor is represented
as the sum of multiple unknown smooth functions and parametric linear predictors
[@woodFastStableRestricted2011]. Although the underlying parameter estimation 
procedure of GAMs are substantially different than WRTDS, both the functional 
form and results have been demonstrated to be similar when assessing nutrient
concentration trends [@beckNumericalQualitativeContrasts2017]. 
Water quality applications of GAMs have included river and catchment 
nutrient concentration and load estimation
[@wangLoadEstimationUncertainties2011;@kroonRiverLoadsSuspended2012;
@kuhnertQuantifyingTotalSuspended2012;@robson_prediction_2015-1;
@hagemannEstimatingNutrientOrganic2016;@mcdowell_implications_2021;
@biagi_novel_2022], and assessment of temporal trends of nutrients
[@beckNumericalQualitativeContrasts2017; @murphyGeneralizedAdditiveModel2019],
phytoplankton [@bergbuschUnexpectedShiftPhytoplankton2021], 
and cyanobacteria [@hayesEffectsLakeWarming2020]. 
Recently GAMs have also been used to link water quality responses in 
receiving water bodies to changes in nonpoint source nutrient inputs 
[@murphyNutrientImprovementsChesapeake2022]. For a substantial discussion on the 
differences (and similarities) between GAMs and WRTDS for water quality 
applications readers are referred to @beckNumericalQualitativeContrasts2017.

@@

## Location and Data {.unnumbered}
Lavaca Bay is a 190 km^2^ estuary with the majority of freshwater inflow provided by the 
Lavaca and Navidad River systems (Fig.\ \ref{fig:fig1}). The Garcitas-Arenosa, 
Placedo Creek, and Cox Bay subwatersheds provide additional freshwater inflows. 
The entire watershed area is 8,149 km^2^ and primarily rural. Watershed 
land cover and land use is 50% grazed pasture and rangeland, 20% cultivated 
cropland (primarily rows crops such as corn, cotton, and sorghum), and 5% 
suburban/urban. Pasture and rangeland is concentrated in the Lavaca River 
watershed, while cultivated crops are generally located along the eastern 
tributaries of the Navidad river. The Lavaca and Navidad River watersheds are a 
combined 5,966 km^2^, or approximately 73% of the entire Lavaca Bay watershed 
area. Discharge from the Navidad River is regulated by Lake Texana which has 
been in operation since 1980. Lake Texana provides 0.210 km^3^ of water 
storage and discharges into the tidal section of the Navidad River which 
ultimately joins the tidal section of the Lavaca River 15 km upstream of the 
confluence with the Lavaca Bay.

```{r fig1, fig.width=6.75, fig.width=5.2, out.width="5.2in", fig.cap="Map of Lavaca Bay and the contribution watershed. The freshwater sites are the most downstream freshwater stream locations with water quality and streamflow data used for nutrient load models. Water quality concentration data at the three Lavaca Bay sites were used to assess relationships between freshwater flows, loads and estuary water quality."}

knitr::include_graphics("Schramm-2023-06-PeerJ_files/figure-latex/fig1.png")

@@

Daily discharges for the Lavaca River (USGS-08164000, Fig.\ \ref{fig:fig1}) 
were obtained from the United States Geologic Survey (USGS) National Water 
Information System using the *dataRetrieval* R package 
[@deciccoDataRetrievalPackagesDiscovering2022]. Gaged daily discharges from 
the outlet of Lake Texana on the Navidad River (USGS-0816425) were provided by 
the Texas Water Development Board (TWDB) 
(April 21, 2022 email from R. Neupane, TWDB).

@@

datasummary(`Station ID` * (`TP (mg/L)` + `NO\\textsubscript{3} (mg/L)` + `Mean Daily Streamflow (cfs)`) ~
               Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
            data = df, output = "data.frame") |> 
  mutate(`Station ID` = c("USGS-08164000", "USGS-08164000", "USGS-08164000",
                          "USGS-08164525", "USGS-08164525", "USGS-08164525")) |> 
  rename("Parameter" = " ") |> 
  kbl(format = "latex", booktabs = TRUE, escape = FALSE,
      caption = "Summary of gauged streamflow and freshwater water quality samples between January 1, 2000 and December 31, 2020.") |>
  collapse_rows(columns = 1, latex_hline = "major")

@@

modelsummary::datasummary(`Station ID` * (`TP (mg/L)`+
                           `NO\\textsubscript{x} (mg/L)` +
                           `TKN (mg/L)` +
                           `Chlorophyll-\\textit{a} (ug/L)` +
                           `DO (mg/L)`) ~
                           Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
                         data = df, output = "data.frame") |> 
  mutate(`Station ID` = c(rep("TCEQ-13383", 5), rep("TCEQ-13384", 5),
                          rep("TCEQ-13563", 5))) |> 
  rename("Parameter" = " ") |> 
  kbl(format = "latex", booktabs = TRUE, escape = FALSE,
      caption = "Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.") |> 
  collapse_rows(columns = 1, latex_hline = "major")

@@

Estimates of NO~3~ and TP loads at the Lavaca River (USGS-08164000) and the 
outlet of Lake Texana on the Navidad River (USGS-08164525) were developed using 
GAMs relating nutrient concentration to river discharge, season, and time.
Separate models were fit at each station for each parameter and used to predict
nutrient concentrations for each day in the study period. GAMs were fit using 
the *mgcv* package in R which makes available multiple 
types of smooth functions with automatic smoothness selection [@woodFastStableRestricted2011]. 
The general form of the model related NO~3~ or TP concentration to a long term
tend, season, streamflow, and two different antecedent discharge terms, 
shown in Eq. \ref{eq1}:

@@

The calculation of model terms for the Lake Texana site were modified
because daily loads are not a function of natural stream flow processes alone,
but of dam releases and nutrient concentrations at the discharge point of the 
lake. *Q*, *ma*, and *fa* terms were calculated based on total gaged inflow from
the 4 major tributaries to the lake. Thin-plate regression splines were used 
for *ddate*, *log1p(Q)*, *fa*, and *ma*. A cyclic cubic regression spline was 
used for *yday* to ensure the ends of the spline match 
(day 1 and day 366 are expected to match). First order penalties were applied 
to the smooths of flow-based variables which penalize departures from a 
flat function to help constrain extrapolations for high flow measurements.

@@

Some prior studies have generated estimates of mean annual TP yields in the 
Lavaca River watershed [Table\ \ref{tab:table5}, @dunnTrendsNutrientInflows1996; 
@rebichSourcesDeliveryNutrients2011; @omaniEstimationSedimentNutrient2014;
@wise_spatially_2019]. Although these studies differ in time periods and 
methodologies, they provide a sanity check for the reasonableness of the 
annual estimates generated in the current study. In a regional assessment of
nutrient loading in river's along the Gulf of Mexico, @dunnTrendsNutrientInflows1996
used the LOADEST model to develop an estimated mean annual yield of 28.9 kg/km^2^.
LOADEST is a multiple linear regression model that fits log transformed pollutant concentrations
to long term, seasonal, and flow based predictors and includes methods for bias correction
when exponentiating the response variable. While the annual yield estimated in 
@dunnTrendsNutrientInflows1996 is lower than the estimate of 35.2 kg/km^2^
in this study, it is important to note that both hydrology and land use was
different for a proportion of the study period. For almost the first ten years of the study period
in @dunnTrendsNutrientInflows1996, the Navidad River was still undammed. More recently
the eastern portion of the watershed has seen large conversions of rice fields to
more traditional row crop production [@kulatHolisticWaterEnergyFoodNexus2019]. 
The apparent increase in mean TP yield might be suggestive of wetter recent conditions.
To evaluate long-term trends in discharge, we fit a GAM relating log-transformed daily
discharges on the Lavaca and Navidad Rivers to season and time (Fig.\ \ref{fig:fig8}).
The long-term trends in discharge indicate watershed discharges were at or above average
from 1972 though the early and mid-1980s. In comparison watershed discharges
since the mid-2000s are at or below average. The trend in mean streamflow alone
is counter to expectations based on changes in nutrient yeild estimates. Studies of
long-term precipitation patterns indicate both the total amount [@dixonTrendDetectionTexas2011] 
and variability of rainfall [@mishraChangesExtremePrecipitation2010;
@fagnantCharacterizingSpatiotemporalTrends2020] has increased in the region 
which provide plausible reasons for increased runoff driven nutrient loading.
Further assessments of the frequency and occurrence of periodic high flow events 
that mobilize sediments and are responsible for the bulk of loading volume would
provide further insight to this discrepancy.

```{r fig8, fig.width=6.85, fig.asp=0.75, fig.cap="Measured daily discharges (log-transformed) and smoothed long-term trends for the Lavaca and Navidad Rivers form 1972 though 2001."}

knitr::include_graphics("Schramm-2023-06-PeerJ_files/figure-latex/fig8.png")

>>>>

---
title: Linking watershed nutrient loading to estuary water quality with semiparametric models
preprint: false
author: 
  - name: Michael Schramm
    affiliation: 1
    corresponding: true
    email: michael.schramm@ag.tamu.edu
affiliation:
  - code: 1
    address: Texas Water Resources Institute, Texas A\&M AgriLife Research, College Station, Texas, United States
abstract: >
  Evaluating estuary water quality responses to reductions (or increases) in nutrient loading attributed to on the ground management actions can be challenging due to the strong influence of environmental drivers on nutrient loads and non-linear relationships. This study applied semiparametric approaches to calculate watershed nutrient loads and assess responses in estuary water quality to seasonally-adjusted freshwater inflow and flow-adjusted nutrient loads in Lavaca Bay, Texas. Lavaca Bay is a secondary embayment on the Texas coast displaying early potential for eutrophication and water quality degradation. Use of flow-adjusted nutrient loads allowed the study to evaluate the response in water quality to changes in nutrient loads driven by anthropogenic sources. Cross-validation indicated that, despite data constraints, semiparametric models performed well at nutrient load prediction. Based on these models, delivered annual nutrient loads varied substantially from year to year. In contrast, minimal changes in flow-normalized loads indicate that nutrient loadings were driven by natural variation in precipitation and runoff as opposed to changes in management of nonpoint sources. Models indicated no evidence of long-term changes in dissolved oxygen or chlorophyll-*a* within Lavaca Bay. However, site specific long-term increases in both organic and inorganic nitrogen are concerning for their potential to fuel eutrophication. Further analysis found freshwater inflow had strong influences on nutrient and chlorophyll-*a* concentrations but there was no evidence that changes in watershed nutrient loading explained additional variation in dissolved oxygen or chlorophyll-*a* concentrations. In addition to providing a baseline assessment of watershed nutrient loading and water quality responses in the Lavaca Bay watershed, this study provides methodological support for the use of semiparametric models in load regression models and estuary assessments.
bibliography: bibliography.bib
output:
  rticles::peerj_article:
    citation_package: biblatex
biblio-style: apa
header-includes:
    - \usepackage{csquotes}
    - \usepackage{booktabs}
    - \usepackage{longtable}
    - \usepackage{array}
    - \usepackage{multirow}
    - \usepackage{wrapfig}
    - \usepackage{float}
    - \usepackage{colortbl}
    - \usepackage{pdflscape}
    - \usepackage{tabu}
    - \usepackage{threeparttable}
    - \usepackage{threeparttablex}
    - \usepackage[normalem]{ulem}
    - \usepackage{makecell}

@@

options(knitr.kable.NA = '')
store <- "C:/Data-Analysis-Projects/lavaca-nutrients/_targets"
```

@@

While WRTDS is a statistical approach developed specifically for water quality applications, 
GAMs are a broadly applicable statistical method. GAMs are a semiparametric 
extension of generalized linear models where the response variable is modeled
as the sum of multiple unknown smooth functions and parametric linear predictors
[@woodFastStableRestricted2011]. Although the underlying parameter estimation 
procedure of GAMs are substantially different than WRTDS, both the functional 
form and results have been demonstrated to be similar when assessing nutrient
concentration trends [@beckNumericalQualitativeContrasts2017]. 
Water quality applications of GAMs have included river and catchment 
nutrient concentration and load estimation
[@wangLoadEstimationUncertainties2011;@kroonRiverLoadsSuspended2012;
@kuhnertQuantifyingTotalSuspended2012;@robson_prediction_2015-1;
@hagemannEstimatingNutrientOrganic2016;@mcdowell_implications_2021;
@biagi_novel_2022], and assessment of temporal trends of nutrients
[@beckNumericalQualitativeContrasts2017; @murphyGeneralizedAdditiveModel2019],
phytoplankton [@bergbuschUnexpectedShiftPhytoplankton2021], 
and cyanobacteria [@hayesEffectsLakeWarming2020]. 
Recently GAMs have also been used to link water quality responses in 
receiving water bodies to changes in nonpoint source nutrient inputs 
[@murphyNutrientImprovementsChesapeake2022]. For a substantial discussion on the 
differences (and similarities) between GAMs and WRTDS for water quality 
applications readers are referred to @beckNumericalQualitativeContrasts2017.

@@

## Location and Data {.unnumbered}
Lavaca Bay is a 190 km^2^ estuary with the majority of freshwater inflow provided by the 
Lavaca and Navidad River systems (Fig.\ \ref{fig:fig1}). Lavaca Bay is relatively
shallow with an average depth of 1.2 m and a generally well mixed and turbid water column 
[@beserespollackLongtermTrendsResponse2011; @montagnaAssessmentRelationshipFreshwater2020].
The Garcitas-Arenosa, Placedo Creek, and Cox Bay subwatersheds provide additional 
freshwater inflows. The proximity to freshwater inputs in the upper Lavaca Bay
results in a strong mean salinity and variance gradient 
[@montagnaAssessmentRelationshipFreshwater2020]. Average salinity ranges from 
12 psu near the Lavaca River mouth to 23 psu at the mouth of Lavaca Bay. The
variance is inversely correlated with mean salinity.
The entire watershed area is 8,149 km^2^ and primarily rural. Watershed 
land cover and land use is 50% grazed pasture and rangeland, 20% cultivated 
cropland (primarily rows crops such as corn, cotton, and sorghum), and 5% 
suburban/urban. Pasture and rangeland is concentrated in the Lavaca River 
watershed, while cultivated crops are generally located along the eastern 
tributaries of the Navidad river. The Lavaca and Navidad River watersheds are a 
combined 5,966 km^2^, or approximately 73% of the entire Lavaca Bay watershed 
area. Discharge from the Navidad River is regulated by Lake Texana which has 
been in operation since 1980. Lake Texana provides 0.210 km^3^ of water 
storage and discharges into the tidal section of the Navidad River which 
ultimately joins the tidal section of the Lavaca River 15 km upstream of the 
confluence with the Lavaca Bay.

```{r fig1, fig.width=6.75, fig.width=5.2, out.width="5.2in", fig.cap="Map of Lavaca Bay watershed. The freshwater sites are the most downstream freshwater stream locations with water quality and streamflow data used for nutrient load models. Water quality concentration data at the three Lavaca Bay sites were used to assess relationships between freshwater flows, loads and estuary water quality."}

knitr::include_graphics("Schramm-2023-08-PeerJ_files/figure-latex/fig1.png")

@@

Daily discharges for the Lavaca River (USGS-08164000, Fig.\ \ref{fig:fig1}) 
were obtained from the United States Geologic Survey (USGS) National Water 
Information System (https://waterdata.usgs.gov/nwis) using the *dataRetrieval* R package 
[@deciccoDataRetrievalPackagesDiscovering2022]. Gaged daily discharges from 
the outlet of Lake Texana on the Navidad River (USGS-0816425) were provided by 
the Texas Water Development Board (TWDB) 
(April 21, 2022 email from R. Neupane, TWDB).

@@

datasummary(`Station ID` * (`TP (mg/L)` + `NO\\textsubscript{3} (mg/L)` + `Mean Daily Streamflow (cfs)`) ~
               Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
            data = df, output = "data.frame") |> 
  mutate(`Station ID` = c("USGS-08164000, Lavaca River", "USGS-08164000, Lavaca River", "USGS-08164000, Lavaca River",
                          "USGS-08164525, Navidad River", "USGS-08164525, Navidad River", "USGS-08164525, Navidad River")) |> 
  rename("Parameter" = " ",
         "Station ID and Watershed" = "Station ID") |> 
  mutate(Method = c("EPA 365.1", "EPA 353.2", "", "EPA 365.1", "EPA 353.2", ""),
         `AWRL\\textsuperscript{a}`= c(0.06, 0.05, NA, 0.06, 0.05, NA),
         `Criteria\\textsuperscript{b}` = c(0.69,1.95,NA,0.2,0.37,NA)) |> 
  kbl(format = "latex", booktabs = TRUE, escape = FALSE,
      caption = "Summary of gauged streamflow and freshwater water quality samples between January 1, 2000 and December 31, 2020.") |>
  column_spec(1:2, width = "2.5cm") |>
  collapse_rows(columns = 1, latex_hline = "major") |> 
  footnote(alphabet = c("Ambient water reporting limit (AWRL) represents the highest concentration that can be used as a reporting limit by a laboratory for inclusion in the state's monitoring data.",
                        "The state of Texas has not adopted numeric criteria for nutrients, these criteria represent screening levels used by the state for assessment purposes."),
           threeparttable = TRUE)

@@

modelsummary::datasummary(`Station ID` * (`TP (mg/L)`+
                           `NO\\textsubscript{x} (mg/L)` +
                           `TKN (mg/L)` +
                           `Chlorophyll-\\textit{a} (ug/L)` +
                           `DO (mg/L)`) ~
                           Mean*Arguments(na.rm=TRUE) + SD*Arguments(na.rm=TRUE) + N,
                         data = df, output = "data.frame") |> 
  mutate(`Station ID` = c(rep("TCEQ-13383", 5), rep("TCEQ-13384", 5),
                          rep("TCEQ-13563", 5))) |> 
  rename("Parameter" = " ") |> 
  mutate(Method = rep(c("EPA 365.1", "EPA 353.2", "EPA 351.2", "EPA 445.0", "ASTM D888-09(C) and TCEQ SOP V1"),3),
         `AWRL\\textsuperscript{a}`= rep(c(0.06, 0.05, 0.2, 3, NA),3),
         `Criteria\\textsuperscript{b}` = rep(c(0.21,NA,NA,11.6,5),3)) |> 
  kbl(format = "latex", booktabs = TRUE, escape = FALSE,
      caption = "Summary of estuary water quality samples collected between January 1, 2005 and December 31, 2020.") |> 
  column_spec(6, width = "3cm") |> 
  collapse_rows(columns = 1, latex_hline = "major") |> 
  footnote(alphabet = c("Ambient water reporting limit (AWRL) represents the highest concentration that can be used as a reporting limit by a laboratory for inclusion in the state's monitoring data.",
                        "The state of Texas has not adopted numeric criteria for nutrients, these criteria represent screening levels used by the state for assessment purposes. NO\\\\textsubscript{x} and TKN parameters are not currently screened by the state and do not have an associated screening level."),
           threeparttable = T,
           escape = F)

@@

Estimates of NO~3~ and TP loads at the Lavaca River (USGS-08164000) and the 
outlet of Lake Texana on the Navidad River (USGS-08164525) were developed using 
GAMs relating nutrient concentration to river discharge, season, and time.
Separate models were fit at each station for each parameter and used to predict
nutrient concentrations for each day in the study period. GAMs were fit using 
the *mgcv* package in R which makes available multiple 
types of smooth functions with automatic smoothness selection [@woodFastStableRestricted2011]. 
The general form of the model related NO~3~ or TP concentration to a long term
trend, season, streamflow, and two different antecedent discharge terms, 
shown in Eq. \ref{eq1}:

@@

The calculation of model terms for the Lake Texana site were modified
because daily loads are not a function of natural stream flow processes alone,
but of dam releases and nutrient concentrations at the discharge point of the 
lake. *Q*, *ma*, and *fa* terms were calculated based on total gaged inflow from
the 4 major tributaries to the lake. Thin-plate regression splines were used 
for *ddate*, *log(1+Q)*, *fa*, and *ma*. A cyclic cubic regression spline was 
used for *yday* to ensure the ends of the spline match 
(day 1 and day 366 are expected to match). First order penalties were applied 
to the smooths of flow-based variables which penalize departures from a 
flat function to help constrain extrapolations for high flow measurements.

@@

Some prior studies have generated estimates of mean annual TP yields in the 
Lavaca River watershed [Table\ \ref{tab:table5}, @dunnTrendsNutrientInflows1996; 
@rebichSourcesDeliveryNutrients2011; @omaniEstimationSedimentNutrient2014;
@wise_spatially_2019]. Although these studies differ in time periods and 
methodologies, they provide a sanity check for the reasonableness of the 
annual estimates generated in the current study. In a regional assessment of
nutrient loading in river's along the Gulf of Mexico, @dunnTrendsNutrientInflows1996
used the LOADEST model to develop an estimated mean annual yield of 28.9 kg/km^2^.
LOADEST is a multiple linear regression model that fits log transformed pollutant concentrations
to long term, seasonal, and flow based predictors and includes methods for bias correction
when exponentiating the response variable. @rebichSourcesDeliveryNutrients2011 
and @wise_spatially_2019 used SPARROW to provide a more recent assessment of 
regional catchment based loadings to the Gulf of Mexico (Table\ \ref{tab:table5}).
SPARROW is a hybrid statistical-process model with the underlying nutrient
load estimation methods based on the previously described LOADEST 
[@schwarzSPARROWSurfaceWaterQuality2006]. The functional form of the LOADEST
regression model is similar to the terms applied in the GAMs used in the current
study. The only study to apply a mechanistic watershed model (SWAT) to estimate
nutrient loadings in the Lavaca River watershed 
@omaniEstimationSedimentNutrient2014 developed estimated yields (42 kg/km^2^) 
similar to the two SPARROW models. Although direct comparisons are complicated
by varying time periods, the estimates in this study do fall within the range of
of previously developed estimates. 
To evaluate changes in long-term trends in discharge might be associated with
the nutrient yield estimates covering different time periods, 
we fit a GAM relating log-transformed daily discharges on the Lavaca River to 
season and time (Fig.\ \ref{fig:fig8}).
The long-term trends in discharge indicate watershed discharges were at or above average
from 1972 though the early and mid-1980s. In comparison watershed discharges
since the mid-2000s are at or below average. 
It is probable that the lower than average discharges observed from 2010 
through 2021 (Fig.\ \ref{fig:fig8}) bias our estimates downward compared to studies
that included higher than average streamflow periods (1995-2005). Overall, the 
ranges of estimated yields among different studies along with the apparent 
large variability in streamflow driven loadings (Fig.\ \ref{fig:fig3}, 
Fig. \ref{fig:fig4}) suggest that the current estimates
of TP loading are reasonable.

```{r fig8, out.width="70%", fig.cap="Measured daily discharges (log-transformed) and smoothed long-term trends for the Lavaca River form 1972 though 2001."}

knitr::include_graphics("Schramm-2023-08-PeerJ_files/figure-latex/fig8.pdf")
